Measuring Directionality in Double Beta Decay and Neutrino Interactions with Kiloton-Scale Scintillation Detectors

### 2013-07-22 Direction_v5.tex
Commit of version 5. Version 5 is the final version for proofreading. --Christoph

%
\author{C.~Aberle}\ucla
\author{A.~Elagin}\chicago
\author{M.~Wetstein}\chicago
\author{H.~J.~Frisch}\chicago
\author{L.~Winslow}\ucla

@@

\pacs{23.40.$-$s, 21.10.Tg, 14.60.Pq, 27.60.$+$j}

@@

\section{Introduction}
Liquid scintillator based detectors are responsible for several of the
critical measurements that have determined our present understanding
of neutrino masses and mixings. These measurements include KamLAND's
measurement of reactor anti-neutrino oscillation at a distance of
$\sim$200~km\cite{kam2013}, Borexino's measurement of $^{7}$Be solar
neutrino oscillation\cite{borexino}, and most recently the short
baseline reactor anti-neutrino experiments that measured oscillations
due to $\theta_{13}$ at a distance of 1~km: Daya Bay\cite{dbtwo},
Double Chooz\cite{dctwo, dchydrogen}, and RENO\cite{reno}.
Scintillator-based neutrino detectors will continue to be important for the
next set of neutrino measurements, from the determination of the
neutrino mass hierarchy\cite{juno,reno50} to elastic scattering
measurements\cite{isodarscatt} and sterile neutrino
searches\cite{isodar,nist}, and for non-proliferation
applications\cite{nucifer, songs}.

The scalability of these detectors to large volumes also makes them
highly competitive for neutrino-less double-beta ($0\nu\beta\beta$)
decay searches in which the final state consists of a pair of
electrons with energies in the few MeV range.  The observation of this
rare decay would prove that the neutrino is a Majorana particle, which
has profound consequences to our understanding of the generation of
mass and may provide a possible explanation of the matter-antimatter
asymmetry in the universe.  Currently one of the best limits for the
$0\nu\beta\beta$ half-life comes from the scintillating detector
KamLAND-Zen\cite{KZ0nu}.

@@

In a liquid-scintillation-based detector, Cherenkov light is also
produced, although most is absorbed and re-emitted as part of the
scintillation processes.  However, some fraction retains its
directional information. If this directional Cherenkov light can be
isolated from the copious isotropic scintillation light, it may be
possible to reconstruct the direction of the primary particle or, in
the case of double beta decay, to determine the existence and topology
of the pair.  The addition of directionality is thus a powerful tool
for background rejection.  In this paper, we develop a technique for
separating the Cherenkov and scintillation light using the photon arrival
times and evaluate several detector technologies that would allow
the realization of direction reconstruction in kilo-ton scale
scintillating neutrino detectors.

\section{Liquid Scintillator Detectors}
Liquid scintillators are `cocktails' of aromatic hydrocarbons. When
charged particles move through a scintillator, the molecules are
excited, predominantly via the non-localized electrons in the
$\pi$-bonds of the phenyl groups~\cite{birks_book}. Vibrational
and rotational modes of the molecules are turned into heat within
picoseconds through collisions with other molecules.  Within $\sim$10
picoseconds, the $\pi$-electrons de-excite to the first excited state
from higher levels through radiation-less transitions. The first
excited state de-excites through photon emission. There are two
characteristic times for this de-excitation, depending if the singlet
state or the triplet state was excited.  The singlet state will
de-excite within nanoseconds while the triplet state de-excites on the
order of 10's or 100's of nanoseconds. These two processes are
fluorescence and phosphorescence respectively. The exact time
constants for these processes are determined by the composition of the
scintillator.

The absorption and emission spectra overlap at some level
in all molecules. Consequently, if there is only one type of molecule in
the scintillator cocktail the light output is reduced due to
inefficiencies in the energy transfer through multiple absorption and
re-emission processes. Aromatic solutes or fluors are added to the
primary solvent to shift the wavelength of the photons to longer
wavelengths where the scintillator is more transparent. This
wavelength-shifting is also used to match the quantum efficiency as a
function of wavelength for the photodetectors being used. One typical
scintillator mixture uses pseudo-cumene as the solvent with 1-5~g/L of
PPO as the fluor. This mixture has a peak emission at about 400~nm
where bialkali photomultiplier tubes (PMTs) are most sensitive and the
pseudo-cumene is relatively transparent.

A good liquid scintillator will produce $\sim$10,000 photons
isotropically per MeV of deposited energy. Although less abundant,
Cherenkov light will be produced as well if a particle is moving
faster than the speed of light in the medium.  This light is emitted
in a cone centered on the direction of the particle trajectory, and
with a continuous spectrum weighted toward shorter wavelengths but
extending well into the red. The spectrum is described
by~\cite{Cherenkov34}:
\begin{equation}
\label{eqCherenkov}
\frac{dN}{d\lambda dx} = \frac{2 \pi \alpha Z^2}{\lambda^2} \left [ 1 - \frac{1}{\beta^2 n(\lambda)^2} \right ]
\end{equation}
where $n(\lambda)$ is the wavelength-dependent index of refraction and
$\beta$ is the velocity of the incoming particle. In a large detector,
the Cherenkov light produced at wavelengths shorter than the
absorption cutoff of the scintillator will be absorbed and re-emitted
as isotropic light, but wavelengths longer than this cutoff will
propagate across the detector, retaining their directional
information. The yield is roughly 60 photons per MeV, assuming a
400~nm absorption cutoff\cite{qdot}. These undisturbed Cherenkov
photons will have timing determined by the group velocity
\cite{group_velocity_article} in the liquid,
\begin{equation}
\label{eqGroup}
v_{g}(\lambda) = \frac{c_{vacuum}}{n(\lambda) - dn(\lambda)/d\textnormal{log}(\lambda)}.
\end{equation}
The longer wavelength Cherenkov photons typically arrive before the
scintillation light, which is slowed by both the scintillation
processes and the shorter wavelengths involved. Thus, with sufficient
timing resolution and sensitivity to longer wavelengths it should be
possible to separate the directional Cherenkov light and the isotropic
scintillation light, and then to reconstruct the direction of the
initial particle.

In $0\nu\beta\beta$, the electrons emerge with a combined energy equal
to the Q-value of the particular isotope. The individual
electrons come out following distributions of energies and angular correlations, 
a probable case being equal division of energy between back-to-back
electrons\cite{phasespace}. This case is shown in FIG.~\ref{detector_view}
for an example $^{116}$Cd $0\nu\beta\beta$ event. Since the decay
half-life is inversely proportional to the phase-space factor, isotopes with
higher Q-values are preferred and due to backgrounds from the daughters
of the $^{238}$U and $^{232}$Th decay chains those with Q-values at or
above 2.6~MeV are most often considered for $0\nu\beta\beta$ searches.
There are hundreds of candidate $0\nu\beta\beta$ isotopes\cite{tabledbb},
but only a handful with large Q-values.  Most of the high Q-value candidates
have been considered as a dopant for a liquid scintillator:
$^{150}$Nd(Q=3.367~MeV)\cite{minfang,nd1}, $^{96}$Zr(Q=3.350~MeV)\cite{zr1},
$^{100}$Mo(Q=3.034~MeV)\cite{mo1}, $^{82}$Se(Q=3.350~MeV)\cite{qdot},
$^{116}$Cd(Q=2.802~MeV)\cite{qdot, cd1}, $^{130}$Te(Q=2.533~MeV)\cite{qdot, biller},
$^{136}$Xe(Q=2.479~MeV)\cite{KZ0nu} and $^{124}$Sn(Q=2.29~MeV)\cite{sn1}.
Xenon gas readily dissolves into liquid scintillator. For the other isotopes,
suitable organometallic compound needs to be found that produces a stable
scintillator with a long attenuation length in the wavelength region of
interest. Recently, nanocrystals called quantum dots have been explored
as an alternative method for doping scintillator\cite{qdot,qdot2}. Quantum dots
have interesting optical properties which are discussed further
in Section~\ref{scintillator_emission_sec}.

The reconstruction of the direction of the two simultaneous
emitted $\sim$1.5~MeV electrons from the $0\nu\beta\beta$ decay is
quite challenging. In this paper, we decided to start our directional
studies with 5~MeV single electrons since reconstruction of direction
in this energy range had never been achieved. This is a logical
energy to use since this is where historically scintillating detectors
become more practical than water Cherenkov detectors. This is also
an energy that is relevant for studies of $^{8}$B solar neutrinos
and neutrino-electron elastic scattering measurement.

\section{Geant4 Simulation}
In order to study the effects relevant to directional reconstruction
in liquid scintillators, a
\GEANT~\cite{geant4one,geant4two} simulation has been constructed. The 
simulation uses \GEANT~Version 4.9.6 with the default liquid scintillator optical model, in which optical photons are
assigned the group velocity in the wavelength region of normal
dispersion.

@@

The detector geometry is a sphere of 6.5~m radius filled with
scintillator. FIG.~\ref{detector_view} shows the geometry and the
Cherenkov light from an example $^{116}$Cd $0\nu\beta\beta$ event. The
default scintillator properties have been chosen to match the KamLAND
scintillator \cite{kamland2003}: 80\% n-Dodecane, 20\% Pseudocumene
(1,2,4-Trimethylbenzene) and 1.52~g/l PPO (2,5-Diphenyloxazole). The
scintillator properties implemented in the simulation include the
atomic composition and density ($\rho$ = 0.78~g/ml), the
wavelength-dependent attenuation length \cite{tajimaMaster} and
refractive index \cite{OlegThesis}, the scintillation emission
spectrum\cite{tajimaMaster}, emission rise time ($\tau_r$ = 1.0~ns)
and emission decay time constants ($\tau_{d1}$ = 6.9~ns and
$\tau_{d2}$ = 8.8~ns with relative weights of 0.87 and 0.13)
\cite{tajimaThesis}, scintillator light yield (LY, 9030 photons/MeV)
and the Birks constant ($kB$ $\approx$ 0.1~mm/MeV)
\cite{ChrisThesis}. Variations from the baseline KamLAND case are
discussed below. Re-emission of absorbed photons in the scintillator
bulk volume and scattering have not yet been included, but are not
expected to change the conclusions here.

The inner sphere surface is used as the photodetector. It is treated
as fully absorbing (no reflections), with a photodetector coverage of
100\%. Two important photodetector properties have been varied: 1)
the transit time spread (TTS, default $\sigma$ = 0.1~ns) and 2) the
wavelength-dependent quantum efficiency (QE) for photoelectron
production. The default is the QE of a bialkali photocathode (Hamamatsu
R7081 PMT)\cite{Hamamatsu_R7081}, for which digitized values come from the Double Chooz \cite{dctwo}
Monte Carlo simulation. We note that the KamLAND 17-inch PMTs use the
same photocathode type with similar quantum efficiency.

\begin{figure*}[tbh]
\begin{center}
        \subfigure[ ~Default
        simulation.]{\includegraphics[scale=0.295]{graphs/6p5Meter_5MeVElectrons_Bialkali_KamlandScintSpec_TIME.pdf}}
        \subfigure[ ~Increased TTS
        (1.28~ns).]{\includegraphics[scale=0.295]{graphs/6p5Meter_5MeVElectrons_Bialkali_KamlandScintSpec_1p28nsTTS_TIME.pdf}}
        \subfigure[ ~Red-sensitive
        photocathode.]{\includegraphics[scale=0.295]{graphs/6p5Meter_5MeVElectrons_RedSensitiveQE_KamlandScintSpec_TIME.pdf}}
        \caption[]{Photo-electron (PE) arrival times after application
        of the transit-time spread (TTS) for the simulation
        of 1000 electrons (5~MeV) with different values of the TTS and
        wavelength response. PEs from
        Cherenkov light (black, solid line) and scintillation light
        (red, dotted line) are
        compared. The dashed vertical line illustrates a time cut at
        34~ns. (a) Default simulation: bialkali photocathode and TTS =
        0.1~ns ($\sigma$). After the 34.0~ns time cut we get 171~PEs
        from scintillation and 108~PEs from Cherenkov light. (b)
        Default simulation settings except for TTS = 1.28~ns (KamLAND
        17 in. PMTs). After the 34.0~ns time cut we get 349~PEs from
        scintillation and 88~PEs from Cherenkov light. (c) Default
        simulation settings except for a GaAsP photocathode. After the
        34.0~ns time cut we get 226~PEs from scintillation and 229~PEs
        from Cherenkov light. \label{time_plots_comparison}}
        \end{center}
\end{figure*}

Before the simulation results for different simulation settings are
discussed, we note that we concentrate on 5~MeV single electrons since
this is the energy at which scintillating detectors become more practical
than water Cherenkov detectors.  We now highlight the effects which
contribute to the timing of the scintillator detector system.  First,
the simulated travel time of the initial 5~MeV electron is between
0.10 and 0.15~ns, while the travel distance is about 3~cm. Second, the
scintillation light emission follows a distribution characterized by
scintillator-specific rise and decay times. Before the solutes in
liquid scintillator can emit optical photons, the energy has to be
transferred from the solvent to the solute. The time constant of this
energy transfer accounts for a rise time in scintillation light
emission. Past neutrino experiments were not highly sensitive to the
effect of the scintillation rise time, which is the reason why there
is a lack of accurate numbers. We assume a rise time of 1.0~ns; more
detailed studies are needed in the future. The two time constants used
to describe the falling edge of the scintillator emission time
distribution (quoted above) are values specific to the KamLAND
scintillator. Third, chromatic dispersion turns out to be an important
effect in a 6.5m-radius detector at the level of precision needed for
direction reconstruction.

@@

The measurement of the arrival times of single photoelectrons is
affected by the transit-time spread (TTS) of the photodetectors, a
number which can be different by orders of magnitude depending on the
detector type. The default TTS of 0.1~ns ($\sigma$) is a value which
can be achieved with the large area picosecond photodetectors
(LAPPDs)\cite{LAPPDSum,LAPPDTDR} and possibly hybrid photodetectors
(HPDs)\cite{hpdThesis}; even significantly lower TTS numbers are
realistic with the LAPPD~\cite{RSI_paper,PSEC4_paper,anode_paper}. We
note that uncertainties in the vertex reconstruction.will produce a
similar effect to the smearing due to the TTS.

In Sections \ref{detector_timing_sec} to
\ref{scintillator_emission_sec}, we study the
photoelectron timing for different detector configurations. We focus
on the idea to increase discrimination between Cherenkov and
scintillation light by using improved detector timing. The primary
quantities provided by the \GEANT~simulation are the photoelectron hit
positions and the detection times after the TTS resolution has been
applied. In Section~\ref{reconstruction_sec} these quantities are then
used for event reconstruction.

@@

The ratio of Cherenkov-induced to scintillation-induced photo-electrons
in the early time window (\RCS) is a useful figure-of-merit when
comparing different simulation settings, since a higher ratio means
more directional information per PE. For the default simulation
settings $R_{C/S}=0.63$.

\begin{figure}
        \begin{center}
        \includegraphics[scale=0.40]{graphs/cos_psi_34_h.pdf}
        \caption[]{The angular distribution of photoelectron hits
        relative to the original electron direction $\cos(\Psi) =
        x_{hit}/\vec{r}_{hit}$. The sample consists of 1000 events
        with a 5~MeV electron produced at the detector center. Default
        simulation settings are used and both Cherenkov and
        scintillation light are included.} 
        \label{Cherenkov_cone}
        \end{center}
\end{figure}

@@

When the 17-inch KamLAND PMTs \cite{tajimaMaster,kume_1983} (TTS =
1.28~ns) are used in the simulation, the broadening of the time
distributions leads to a strongly decreased ratio of Cherenkov over
scintillation light ($R_{C/S}=0.25$) for $t<34$~ns (see
FIG.~\ref{time_plots_comparison} (b)). This shows that a low
photodetector TTS is critical for directionality reconstruction and
motivates the use of novel detector types.

\section{Detector Wavelength Response}
\label{detector_wavelength_response_sec} 
In addition to decreasing the photodetector TTS to enhance $R_{C/S}$,
it is possible to optimize the wavelength-dependence of the
photocathode. Since Cherenkov photons which passed through meters of
scintillator have longer average wavelengths than scintillation
photons, a photodetector which is more sensitive at long wavelengths
increases not only the absolute number of PEs but also the ratio
between Cherenkov- and scintillation-induced PEs.

We have run the simulation with the QE of an extended red-sensitive
GaAsP photocathode (Hamamatsu R3809U-63)~\cite{Hamamatsu_R3899U}.
FIG.~\ref{time_plots_comparison}(c) shows the results for the
modified simulation with high QE in the red spectral region. The
higher absolute number of photo-electrons coming from Cherenkov light
(factor of $\approx$ 2) and the increased Cherenkov/scintillation
ratio (1.6) in the early time window would significantly improve the
directionality reconstruction.
% Can we quote a value of R for this? hjf

\section{Scintillator Emission Spectrum}
\label{scintillator_emission_sec}
An alternative route towards increasing the separation in time between
Cherenkov and scintillation photon hits is the tuning of the
scintillator emission spectrum. Recently, the use of quantum dots
(QDs) in liquid scintillators has been studied as a possibility to
improve future large scale neutrino experiments~\cite{qdot2}. One
major motivation for quantum-dot-doped scintillator is control of the
emission spectra by tuning the size or composition of the quantum
dots.

The emission spectrum of commercial alloyed core/shell
CdS$_x$Se$_{1-x}$/ZnS quantum dots was measured in
Ref.~\cite{qdot2}. This spectrum shows a symmetric peak centered
around 461~nm with FWHM = 29~nm.  In order to isolate the effect of
the different emission spectrum, the other simulation settings,
including the KamLAND absorption spectrum, were kept unchanged; we
find $R_{C/S}=0.17$ for the default 34-nsec timing cut.  Compared to
the default case shown in FIG.~\ref{time_plots_comparison}(a) the
separation is worse (as expected) because the scintillation light
wavelengths are higher than in the KamLAND emission spectrum.

However, advances in the production of commercial quantum dot samples
could yield quantum dots which have similar, single peak emission
shapes at lower wavelengths. This case has been simulated using the
same spectral shape of the measured core-shell quantum dot emission
but shifted to lower wavelengths such that the emission peak is
centered at 384~nm. This peak emission value has been measured for
other types of QDs, however with a much more pronounced tail
\cite{qdot2}. The resulting PE time distribution shows improved
separation of Cherenkov and scintillation light compared to the
default simulation. After the 34.0~ns cut on the TTS-smeared PE time
we obtain a Cherenkov/scintillation ratio of $R_{C/S}=0.86$ (107 PE
from Cherenkov light and 124 PE from scintillation). The number of
Cherenkov-induced PEs after the time cut is unchanged while the number
of PEs coming from scintillation light is decreased due to the higher
average photon travel times.

>>>>

%
\author{C.~Aberle}\ucla
\author{A.~Elagin}\chicago
\author{H.~J.~Frisch}\chicago
\author{M.~Wetstein}\chicago
\author{L.~Winslow}\ucla

@@

\pacs{23.40.$-$s, 14.60.Pq, 13.15.+g, 29.40.Mc}

@@

\section{Introduction}
Liquid scintillator-based detectors are responsible for several of the
critical measurements that have determined our present understanding
of neutrino masses and mixings. These measurements include KamLAND's
measurement of reactor anti-neutrino oscillation at a distance of
$\sim$200~km\cite{kam2013}, Borexino's measurement of $^{7}$Be solar
neutrino oscillation\cite{borexino}, and most recently the short
baseline reactor anti-neutrino experiments that measured oscillations
due to $\theta_{13}$ at a distance of 1~km: Daya Bay\cite{dbtwo},
Double Chooz\cite{dctwo, dchydrogen}, and RENO\cite{reno}.
Scintillator-based neutrino detectors will continue to be important for the
next set of neutrino measurements, from the determination of the
neutrino mass hierarchy\cite{juno,reno50} to elastic scattering
measurements\cite{isodarscatt} and sterile neutrino
searches\cite{isodar,nist}, and for non-proliferation
applications\cite{nucifer, songs}.

The scalability of these detectors to large volumes also makes them
highly competitive for neutrino-less double-beta ($0\nu\beta\beta$)
decay searches in which the final state consists of a pair of
electrons with energies in the few MeV range.  The observation of this
rare decay would prove that the neutrino is a Majorana particle, which
would have profound consequences to our understanding of the generation of
mass and may provide a possible explanation of the matter-antimatter
asymmetry in the universe.  Currently one of the best limits for the
$0\nu\beta\beta$ half-life comes from the scintillating detector
KamLAND-Zen\cite{KZ0nu}.

@@

In a liquid-scintillation-based detector, Cherenkov light is also
produced, although most is absorbed and re-emitted as part of the
scintillation processes.  However, some fraction retains its
directional information. If this directional Cherenkov light can be
isolated from the copious isotropic scintillation light, it may be
possible to reconstruct the direction of the primary particle or, in
the case of double-beta decay, to determine the existence and topology
of the pair.  The addition of directionality is thus a powerful tool
for background rejection.  In this paper, we develop a technique for
separating the Cherenkov and scintillation light using the photon arrival
times and evaluate several detector advances in timing, photodetector 
spectral response, and scintillator emission spectra that would allow
the realization of direction reconstruction in kilo-ton scale
scintillating neutrino detectors.

\section{Liquid Scintillator Detectors}
Liquid scintillators are `cocktails' of aromatic hydrocarbons. When
charged particles move through a scintillator, the molecules are
excited, predominantly via the non-localized electrons in the
$\pi$-bonds of the phenyl groups\cite{birks_book}. Vibrational
and rotational modes of the molecules are turned into heat within
picoseconds through collisions with other molecules.  Within $\sim$10
picoseconds, the $\pi$-electrons de-excite to the first excited state
from higher levels through radiation-less transitions. The first
excited state can de-excite through photon emission. There are two
characteristic times for this de-excitation, depending if the singlet
state or the triplet state was excited.  The singlet state will
de-excite within nanoseconds while the triplet state de-excites on the
order of 10's or 100's of nanoseconds. These two processes are
fluorescence and phosphorescence respectively. The exact time
constants for these processes are determined by the composition of the
scintillator.

The absorption and emission spectra overlap at some level
in all molecules. Consequently, if there is only one type of molecule in
the scintillator cocktail the light output is reduced due to
inefficiencies in the energy transfer through multiple absorption and
re-emission processes. Aromatic solutes or fluorophores are added to the
primary solvent to shift the wavelengths of the photons to higher values 
where the scintillator is more transparent. This
wavelength-shifting is also used to match the quantum efficiency as a
function of wavelength for the photodetectors being used. One typical
scintillator mixture uses pseudocumene as the solvent with 1-5~g/L of
PPO as the fluorophore. This mixture has a peak emission at about 400~nm
where bialkali photomultiplier tubes (PMTs) are most sensitive and the
pseudocumene is relatively transparent.

A good liquid scintillator will produce $\sim$10,000 photons
isotropically per MeV of deposited energy. Although less abundant,
Cherenkov light will be produced as well if a particle is moving
faster than the speed of light in the medium.  This light is emitted
in a cone centered on the direction of the particle trajectory, and
with a continuous spectrum weighted toward shorter wavelengths but
extending well into the red. The spectrum is described
by\cite{Cherenkov34}:
\begin{equation}
\label{eqCherenkov}
\frac{d^2N}{d\lambda dx} = \frac{2 \pi \alpha Z^2}{\lambda^2} \left [ 1 - \frac{1}{\beta^2 n(\lambda)^2} \right ]
\end{equation}
where $n(\lambda)$ is the wavelength-dependent index of refraction and
$\beta$ is the velocity of the incoming particle. In a large detector,
the Cherenkov light produced at wavelengths shorter than the
absorption cutoff of the scintillator will be absorbed and re-emitted
as isotropic light, but wavelengths longer than this cutoff will
propagate across the detector, retaining their directional
information. The yield is roughly 60 photons per MeV, assuming a
400~nm absorption cutoff\cite{qdot}. These undisturbed Cherenkov
photons will have timing determined by the group 
velocity\cite{group_velocity_article,pdg_review_2012,tamm1939} in the liquid,
\begin{equation}
\label{eqGroup}
v_{g}(\lambda) = \frac{c_{vacuum}}{n(\lambda) - dn(\lambda)/d\textnormal{log}(\lambda)}.
\end{equation}
The longer wavelength Cherenkov photons typically arrive before the
scintillation light, which is slowed by both the scintillation
processes and the shorter wavelengths involved. Thus, with sufficient
timing resolution and sensitivity to longer wavelengths it should be
possible to separate the directional Cherenkov light and the isotropic
scintillation light, and then to reconstruct the direction of the
initial particle.

In $0\nu\beta\beta$, the electrons emerge with a combined energy equal
to the Q-value of the particular isotope. The individual
electrons follow distributions of energies and angular correlations, 
a probable case being equal division of energy between back-to-back
electrons\cite{phasespace}. This case is shown in FIG.~\ref{detector_view}
for an example $^{116}$Cd $0\nu\beta\beta$ event. Since the decay
half-life is inversely proportional to the phase-space factor, isotopes with
higher Q-values are preferred and due to backgrounds from the daughters
of the $^{238}$U and $^{232}$Th decay chains those with Q-values at or
above 2.6~MeV are most often considered for $0\nu\beta\beta$ searches.
There are hundreds of candidate $0\nu\beta\beta$ isotopes\cite{tabledbb},
but only a handful with large Q-values.  Most of the high Q-value candidates
have been considered as a dopant for a liquid scintillator:
$^{150}$Nd(Q=3.367~MeV)\cite{minfang,nd1}, $^{96}$Zr(Q=3.350~MeV)\cite{zr1},
$^{100}$Mo(Q=3.034~MeV)\cite{mo1}, $^{82}$Se(Q=3.350~MeV)\cite{qdot},
$^{116}$Cd(Q=2.802~MeV)\cite{qdot, cd1}, $^{130}$Te(Q=2.533~MeV)\cite{qdot, biller},
$^{136}$Xe(Q=2.479~MeV)\cite{KZ0nu} and $^{124}$Sn(Q=2.29~MeV)\cite{sn1}.
Xenon gas readily dissolves into liquid scintillator. For the other isotopes,
a suitable organometallic compound needs to be found that produces a stable
scintillator with a long attenuation length in the wavelength region of
interest. As an alternative to single-atom-doping, recently nanocrystals 
formed by candidate isotopes have been explored as a dopant\cite{qdot,qdot2}.

\section{Geant4 Simulation}
In order to study the effects relevant to directional reconstruction
in liquid scintillators, a
\GEANT\cite{geant4one,geant4two} simulation has been constructed. The
simulation uses \GEANT~Version 4.9.6 with the default liquid scintillator
optical model, in which optical photons are
assigned the group velocity in the wavelength region of normal
dispersion.

@@

The detector geometry is a sphere of 6.5~m radius filled with
scintillator. FIG.~\ref{detector_view} shows the geometry and the
Cherenkov light from an example $^{116}$Cd $0\nu\beta\beta$ event. The
default scintillator properties have been chosen to match the KamLAND
scintillator\cite{kamland2003}: 80\% n-dodecane, 20\% pseudocumene
(1,2,4-trimethylbenzene) and 1.52~g/l PPO (2,5-diphenyloxazole). The
scintillator properties implemented in the simulation include the
atomic composition and density ($\rho$ = 0.78~g/ml), the
wavelength-dependent attenuation length\cite{tajimaMaster} and
refractive index\cite{OlegThesis}, the scintillation emission
spectrum\cite{tajimaMaster}, emission rise time ($\tau_r$ = 1.0~ns)
and emission decay time constants ($\tau_{d1}$ = 6.9~ns and
$\tau_{d2}$ = 8.8~ns with relative weights of 0.87 and 
0.13)\cite{tajimaThesis}, scintillator light yield (9030 photons/MeV)
and the Birks constant ($kB$ $\approx$ 0.1~mm/MeV)\cite{ChrisThesis}. 
Variations from the baseline KamLAND case are
discussed below. Re-emission of absorbed photons in the scintillator
bulk volume and scattering have not yet been included, but are not
expected to change the conclusions here.

The inner sphere surface is used as the photodetector. It is treated
as fully absorbing (no reflections), with a photodetector coverage of
100\%. Two important photodetector properties have been varied: 1)
the transit time spread (TTS, default $\sigma$ = 0.1~ns) and 2) the
wavelength-dependent quantum efficiency (QE) for photoelectron
production. The default is the QE of a bialkali photocathode (Hamamatsu
R7081 PMT)\cite{Hamamatsu_R7081}, for which digitized values come from the Double Chooz\cite{dctwo}
Monte Carlo simulation. We note that the KamLAND 17-inch PMTs use the
same photocathode type with similar quantum efficiency.

\begin{figure*}[tbh]
\begin{center}
        \subfigure[ ~Default
        simulation.]{\includegraphics[scale=0.295]{graphs/6p5Meter_5MeVElectrons_Bialkali_KamlandScintSpec_TIME.pdf}}
        \subfigure[ ~Increased TTS
        (1.28~ns).]{\includegraphics[scale=0.295]{graphs/6p5Meter_5MeVElectrons_Bialkali_KamlandScintSpec_1p28nsTTS_TIME.pdf}}
        \subfigure[ ~Red-sensitive
        photocathode.]{\includegraphics[scale=0.295]{graphs/6p5Meter_5MeVElectrons_RedSensitiveQE_KamlandScintSpec_TIME.pdf}}
        \caption[]{Photo-electron (PE) arrival times after application
        of the transit-time spread (TTS) for the simulation
        of 1000 electrons (5~MeV) with different values of the TTS and
        wavelength response. PEs from
        Cherenkov light (black, solid line) and scintillation light
        (red, dotted line) are
        compared. The dash-dotted vertical line illustrates a time cut at
        34~ns. (a) Default simulation: bialkali photocathode and TTS =
        0.1~ns ($\sigma$). After the 34.0~ns time cut we get 171~PEs
        from scintillation and 108~PEs from Cherenkov light. (b)
        Default simulation settings except for TTS = 1.28~ns (KamLAND
        17 in. PMTs). After the 34.0~ns time cut we get 349~PEs from
        scintillation and 88~PEs from Cherenkov light. (c) Default
        simulation settings except for a GaAsP photocathode. After the
        34.0~ns time cut we get 226~PEs from scintillation and 229~PEs
        from Cherenkov light. \label{time_plots_comparison}}
        \end{center}
\end{figure*}

The initial studies described here are made with single 5~MeV electrons. 
Future studies will work on increasing the complexity and lowering 
the electron energies as is needed for $0\nu\beta\beta$ studies. Three 
effects primarily contribute to the timing of the scintillator detector 
system. First, the simulated travel time of the initial 5~MeV electron is between
0.10 and 0.15~ns, while the travel distance is about 3~cm. Second, the
scintillation light emission follows a distribution characterized by
scintillator-specific rise and decay times. Before the solutes in
liquid scintillator can emit optical photons, the energy has to be
transferred from the solvent to the solute. The time constant of this
energy transfer accounts for a rise time in scintillation light
emission. Past neutrino experiments were not highly sensitive to the
effect of the scintillation rise time, which is the reason why there
is a lack of accurate numbers. We assume a rise time of 1.0~ns; more
detailed studies are needed in the future. The two time constants used
to describe the falling edge of the scintillator emission time
distribution (quoted above) are values specific to the KamLAND
scintillator. Third, chromatic dispersion turns out to be an important
effect in a 6.5m-radius detector at the level of precision needed for
direction reconstruction.

@@

The measurement of the arrival times of single photoelectrons is
affected by the transit-time spread (TTS) of the photodetectors, a
number which can be different by orders of magnitude depending on the
detector type. The default TTS of 0.1~ns ($\sigma$) is a value which
can be achieved with the large area picosecond photodetectors
(LAPPDs)\cite{LAPPDSum,LAPPDTDR} and possibly hybrid photodetectors
(HPDs)\cite{hpdThesis}; even significantly lower TTS numbers are
realistic with the LAPPD\cite{RSI_paper,PSEC4_paper,anode_paper}. We
note that uncertainties in the vertex reconstruction.will produce a
similar effect to the smearing due to the TTS.

In Sections \ref{detector_timing_sec} to
\ref{scintillator_emission_sec}, we study the
photoelectron timing for different detector configurations. We focus
on the idea for increasing discrimination between Cherenkov and
scintillation light by using improved detector timing. The primary
quantities provided by the \GEANT~simulation are the photoelectron hit
positions and the detection times after the TTS resolution has been
applied. In Section~\ref{reconstruction_sec} these quantities are then
used for event reconstruction.

@@

The ratio of Cherenkov-induced to scintillation-induced photoelectrons
in the early time window ($R_{C/S}$) is a useful figure-of-merit when
comparing different simulation settings, since a higher ratio means
more directional information per PE. For the default simulation
settings $R_{C/S}=0.63$.

\begin{figure}
        \begin{center}
        \includegraphics[scale=0.40]{graphs/cos_psi_34_h.pdf}
        \caption[]{The angular distribution of photoelectron hits
        relative to the original electron direction, $\cos(\Psi) =
        x_{hit}/\vec{r}_{hit}$. The sample consists of 1000 events
        with a 5~MeV electron produced at the detector center. Default
        simulation settings are used and both Cherenkov and
        scintillation light are included.} 
        \label{Cherenkov_cone}
        \end{center}
\end{figure}

@@

When the 17-inch KamLAND PMTs\cite{tajimaMaster,kume_1983} (TTS =
1.28~ns) are used in the simulation, the broadening of the time
distributions leads to a strongly decreased ratio of Cherenkov over
scintillation light ($R_{C/S}=0.25$) for $t<34$~ns (see
FIG.~\ref{time_plots_comparison} (b)). This shows that a low
photodetector TTS is critical for directionality reconstruction and
motivates the use of novel detector types.

\section{Detector Wavelength Response}
\label{detector_wavelength_response_sec} 
In addition to decreasing the photodetector TTS to enhance $R_{C/S}$,
it is possible to optimize the wavelength-dependence of the
photocathode. Since Cherenkov photons which passed through meters of
scintillator have on average longer wavelengths than scintillation
photons, a photodetector which is more sensitive at long wavelengths
increases not only the absolute number of PEs but also the ratio
between Cherenkov- and scintillation-induced PEs.

We have run the simulation with the QE of an extended red-sensitive
GaAsP photocathode (Hamamatsu R3809U-63)\cite{Hamamatsu_R3899U}.
FIG.~\ref{time_plots_comparison}(c) shows the results for the
modified simulation with high QE in the red spectral region. The
higher absolute number of photoelectrons coming from Cherenkov light
(factor of $\approx$ 2) and the increased Cherenkov/scintillation
ratio ($R_{C/S}=1.6$) in the early time window would significantly improve the
directionality reconstruction.

\section{Scintillator Emission Spectrum}
\label{scintillator_emission_sec}
An alternative route towards increasing the separation in time between
Cherenkov and scintillation photon hits is the tuning of the
scintillator emission spectrum. Recently, the use of quantum dots
(QDs) in liquid scintillators has been studied as a possibility to
improve future large scale neutrino experiments\cite{qdot,qdot2}. One
major motivation for quantum-dot-doped scintillator is control of the
emission spectra by tuning the size or composition of the quantum
dots; quantum dots can also provide a mechanism for introducing an 
isotope for studying double-beta decay.

The emission spectrum of commercial alloyed core/shell
CdS$_x$Se$_{1-x}$/ZnS quantum dots was measured in
Ref.\cite{qdot2}. This spectrum shows a symmetric peak centered
around 461~nm with FWHM = 29~nm.  In order to isolate the effect of
the different emission spectrum, the other simulation settings,
including the KamLAND absorption spectrum, were kept unchanged; we
find $R_{C/S}=0.17$ for the default 34~ns timing cut.  Compared to
the default case shown in FIG.~\ref{time_plots_comparison}(a) the
separation is worse (as expected) because the scintillation light
wavelengths are higher than in the KamLAND emission spectrum.

However, advances in the production of commercial quantum dot samples
could yield quantum dots which have similar, single peak emission
shapes at lower wavelengths. This case has been simulated using the
same spectral shape of the measured core-shell quantum dot emission
but shifted to lower wavelengths such that the emission peak is
centered at 384~nm. This peak emission value has been measured for
other types of QDs, however with a much more pronounced 
tail\cite{qdot2}. The resulting PE time distribution shows improved
separation of Cherenkov and scintillation light compared to the
default simulation. After the 34.0~ns cut on the TTS-smeared PE time
we obtain a Cherenkov/scintillation ratio of $R_{C/S}=0.86$ (107 PE
from Cherenkov light and 124 PE from scintillation). The number of
Cherenkov-induced PEs after the time cut is unchanged while the number
of PEs coming from scintillation light is decreased due to the higher
average photon travel times.
